% !TEX root =../LibroTipoETSI.tex
\chapter*{Acknowledgements}
%\pagestyle{especial}
\pagestyle{empty}
%\chaptermark{Agradecimientos}
\phantomsection
%\addcontentsline{toc}{listasf}{Agradecimientos}
%\vspace{1cm}
%{\huge{Agradecimientos}}
%\vspace{1cm}

\lettrine[lraise=-0.1, lines=2, loversize=0.25]{F}{i}rst and foremost, I would like to express my deepest gratitude to my university peers. Throughout these years, their help and support have been fundamental in overcoming difficult moments. I have learned so much from them, and they have made this journey much more manageable and enriching. Thank you for always being willing to collaborate, for the shared moments, and for the laughter that lightened up the days of study.

To my family, especially my parents, I want to thank you from the bottom of my heart. Thank you for giving me an education that has allowed me to become the person I am today. Your love and sacrifices have been the cornerstone for achieving my goals and continuing to strive every day. I will always be grateful for everything you have done for me.

I also want to thank my grandparents, who have been a constant example of effort and hard work. Thank you for always being proud of every step I have taken in my life. Your dedication and life example have been a constant source of inspiration.

To my sister, thank you for being a role model, not only in studies but also in how to enjoy life. Your positive attitude and your ability to find a balance between responsibilities and fun have been a great inspiration to me.

Lastly, I would like to express my gratitude to my internship supervisor, Juan Antonio. Thank you for your valuable guidance and support throughout this project. Your commitment and willingness to assist me every step of the way have been essential to the development of this work. I deeply appreciate your dedication and the confidence you have placed in me.

%gradecemos}, a todos nuestros maestros, cuanto nos enseñaron.

{\flushleft{\hfill \emph{Antonio Luis González Hernández}}}%
\vspace{-.3cm}
{\flushleft{\hfill \emph{Seville, 2024}}}%